Mergeable Types

### 2017-03-15 paper/examples.tex
Fixed many errors.

The \drawsome example demonstrates the \name's utility in building
concurrent applications with conceptual state sharing that make them
suitable for transparent deployment in a distributed setting. The
model allows concurrent computations to be composed around any
mergeable type. We now present a series of examples that demonstrate
how mergeable types are built - either by defining merge functions for
ML data types, or by composing existing mergeable types.

@@

A rope is a data structure that facilitates efficient implementations
of operations such as concatenation, splitting, lookup and insertion
at a random position etc., for a list of values in a distributed
setting. The values are usually characters, with their lists being
strings. The data structure is a binary tree, whose leaf nodes store
substrings of a larger string such that their left-to-right (in-order)
concatenation yields the larger string. The idea is to facilitate
efficient lookups and insertions at random positions in the string by
storing position indexes in the internal nodes of the tree.

Fig.~\ref{fig:mrope-sig} shows a rope interface in OCaml. The type of
the list of values is left abstract (\C{A.t}); it can be instantiated
with any module that supports \C{concat} and \C{split} operations. The
interface shown in Fig.~\ref{fig:mrope-sig} is that of a mergeable
rope (\C{MRope}), whose \C{merge} implementation is shown in
Fig.~\ref{fig:mrope-merge}. The function returns one of its versions
(\C{l} or \C{r}), if the other one is the same as the common
ancestor. Otherwise, it calls \C{merge\_rec}, which recursively
compares the structures of merging ropes, and their common ancestor.
The function recurses until one of the ropes differs from the other
two, or all of them are leaf nodes, at which point it relies on the
(abstract) list merge function (\C{A.merge}) to merge the lists
represented by the ropes. \C{merge\_rec} uses auxiliary functions,
\C{mk\_rope} and \C{create\_index}, whose definitions are not shown to
construct a rope given a list (\C{A.t}), and to compute the index at a
node, given its left sub-tree, resp.

One feature of the mergeable rope implementation that is characteristic
of mergeable types in general is compositionality; a rope is a
polymorphic mergeable data type that composes mergeability over its
polymorphic constituents.   Compositionality is concretized by the
\C{MRope} signature , which requires the type of items in the list
(\C{A.t}) to also be mergeable (\C{A.merge}).

@@

As an abstract data type, lists support three operations: \C{insert x
  i l}, to insert an element \C{x} at position \C{i} in the list
\C{l}, \C{delete i l}, to delete the element at \C{i}'th position, and
\C{subst i x l}, to substitute the element at the \C{i}'th position
with \C{x}.  Lists of mergeable items are mergeable if they preserve
the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike, for example, Amazon's replicated shopping cart~\cite{Dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}

\noindent The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml.  The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The signature
		specified in Fig.~\ref{fig:mlist} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{fig:list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
		have the same effect as it did on $v$. We say \C{D(1)} in $s_1$ has to be
		transformed w.r.t the sequence $s_2 = \C{[I(c,0);S(3,c,s)]}$.  The function
		\C{op\_transform} (shown in the Appendix) computes such
		\emph{operational transformations} for an edit sequence $s_1$ w.r.t another
		edit sequence $s_2$. A notable transformation rule is for a substitution
		w.r.t another substitution at the same position. Since the substituting
		items could be different, the function relies on the item merge function
		(\C{A.merge}) to merge them into a single item. For instance, in
		Fig.~\ref{fig:list-eg}, there are simultaneous substitutions to the item
		\C{c} in $s_1$ and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does
		it with \C{s}. The merged list therefore substitutes \C{c} with the merge
		of \C{d} and \C{s}, defined as per the merge semantics of \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is a shopping list application that allows its users
to collaboratively build a shopping list (listing items in the
decreasing order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}).  The
merge function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
i.e., \C{MList(Item)} (although a grocery shopping list may not be
large enough to justify using ropes, \C{MRope(MList(Item))} is
certainly allowed).  \C{MList(Item).merge} automatically lifts the
merge semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
\begin{wrapfigure}{r}{.35\textwidth}
  \centering
  \includegraphics[width=0.25\textwidth]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{.25in}
\end{wrapfigure}
and Bob's quantity as per the definition of \C{Qty.merge}.
The examples described above make certain assumptions about the
concurrency model and the underlying system, such as the existence of
a single lowest common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

@@

>>>>

The \drawsome example demonstrates \name's utility in building
concurrent applications with conceptual state sharing, which makes
them suitable for transparent deployment in a distributed setting. The
model allows concurrent computations to be composed around any
mergeable type. We now present a series of examples of mergeable types
built either by defining merge functions for ML data types, or by
composing existing mergeable types.

@@

A rope is a data structure that facilitates efficient implementations
of operations such as concatenation, splitting, lookup and insertion
at a random position etc., for a list of values. The values are
usually characters, their lists being strings. A rope is essentially a
binary tree, whose leaf nodes store substrings of a larger string such
that their left-to-right (in-order) concatenation yields the larger
string. The idea is to facilitate efficient lookups and insertions at
random positions in the string by storing position indexes in the
internal nodes of the tree.

Fig.~\ref{fig:mrope-sig} shows a rope interface in OCaml. The type of
the list of values is left abstract (\C{A.t}); it can be instantiated
with any module that supports \C{concat} and \C{split} operations. The
interface shown in Fig.~\ref{fig:mrope-sig} is that of a mergeable
rope (\C{MRope}), whose \C{merge} implementation is shown in
Fig.~\ref{fig:mrope-merge}. The function returns one of its versions
(\C{l} or \C{r}), if the other one is the same as the common ancestor.
Otherwise, it calls \C{merge\_rec}, which recursively compares the
structures of merging ropes, and their common ancestor.  The function
recurses until one of the ropes differs from the other two, or all of
them are leaf nodes, at which point it relies on the (abstract) list
merge function (\C{A.merge}) to merge the lists represented by the
ropes. \C{merge\_rec} uses auxiliary functions, \C{mk\_rope} and
\C{create\_index} (definitions not shown), respectively, to construct
a rope given a list (\C{A.t}), and to compute the index at a node
given its left sub-tree.

A notable feature of the mergeable rope implementation, which is also
a characteristic of mergeable types in general, is compositionality. A
rope is a polymorphic mergeable data type that composes mergeability
over its polymorphic constituents.  Compositionality is concretized by
the \C{MRope} signature, which requires the type of items in the list
(\C{A.t}) to also be mergeable (\C{A.merge}).

@@

As an abstract data type, \C{list} supports three operations:
\C{insert x i l}, to insert an element \C{x} at position \C{i} in the
list \C{l}, \C{delete i l}, to delete the element at \C{i}'th position
in \C{l}, and \C{subst i x l}, to substitute the element at the
\C{i}'th position with \C{x} in \C{l}.  Lists of mergeable items are
mergeable if they preserve the following properties:
\begin{itemize}
  \item Elements present in both the merging lists will be present in
  the final list. No element is deleted unless it is deleted in at
  least one list.
  \item An element deleted in one list will not reappear in the merged
  list (unlike, for example, Amazon's replicated shopping cart~\cite{Dynamo}).
  \item The order of elements is retained. That is, if $x$ occurs
  before $y$ in one of the lists, and if both elements are present in
  the merged list, then $x$ occurs before $y$ in the merged list.
\item If an element $y$ is substituted for an element $x$ in one list,
  then $y$ is also substituted for $x$ in the merged list (i.e., $x$
  is absent and $y$ is present), unless $x$ is deleted or substituted
  in the other list. In the former case, deletion wins, and neither
  $x$ nor $y$ occur in the merged list. In the latter case, if $x$ is
  substituted by a different element $z$ in the other list, then the
  merged list substitutes $x$ by a merge of $y$ and $z$, defined as
  per the merge semantics of the list element type.
\end{itemize}

\noindent The merge operation for lists is composed of two separate functions,
\C{edit\_seq} and \C{op\_transform}. Both these functions have been
implemented in less than 50 lines of OCaml.  The functions are described below.
\begin{itemize}
	\item \C{edit\_seq} takes a pair of lists, \C{v} and $\C{v'}$, and computes
		the shortest sequence of list operations that need to be applied on \C{v}
		to obtain \C{v'}. Such a sequence is called an \emph{edit sequence}. The
		length of the sequence corresponds to the standard notion of the \emph{edit
		distance} between the two lists, which can be computed in polynomial
		time~\cite{wagner-fischer}.  \C{edit\_seq} modifies one such algorithm to
		also compute the edit sequence (an \C{edit list}). The signature
		specified in Fig.~\ref{fig:mlist} represents edits using the type
		\C{edit}.  Constructors \C{I}, \C{D}, and \C{S} stand for \C{insert},
		\C{delete}, and \C{subst}, respectively.\footnote{ Note that \C{I}, \C{D}
		and \C{S} are \emph{not} effects; no list function generates them, and
		they are not exchanged between concurrent threads. They are simply tags
		used by the list merge operation for convenience.  Indeed, to better
		appreciate the virtues of mergeable types, we encourage the reader to
		think about how one might implement a replicated list with this
		functionality using an explicit effect representation as described in
		Sec.~\ref{sec:intro}.}  The subst constructor also carries the \C{A.t}
		element that was substituted. The element serves as the \C{lca} argument to
		\C{A.merge} in case of concurrent conflicting substitutions.
		Fig.~\ref{fig:list-eg} illustrates edit sequences for a sample list.  The
		sequence \C{[I(c,0); S(3,c,s)]} maps the list \C{[a;b;c]} to \C{[c;a;b;s]}
		because \C{subst 3 s (insert c 0 [a;b;c]) = [c;a;b;s]}.
	\item  \C{op\_transform} takes a pair of edit sequences, $s_1$ and $s_2$,
		that map a list $v$ to two different lists, $v_1$ and $v_2$ (e.g.,
		Fig.~\ref{fig:list-eg}), and transforms $s_1$ to $s_1'$ such that $s_1'$
		has the same effect on $v_2$ as $s_1$ had on $v$.  For instance, in
		Fig.~\ref{fig:list-eg}, let $s_1$ be the edit sequence \C{[D(1); S(1,c,d)]},
		which maps the list \C{[a;b;c]} ($v$) to \C{[a;d]} ($v_1$). The \C{D} edit
		deletes the first element of $v$. However, if $s_1$ is to be applied on the
		list \C{[c;a;b;s]} ($v_2$), which has a new element at position 0, the
		\C{D} edit must delete the second element from $v_2$, if it were to
    have the same effect as it did on $v$. We say \C{D(1)} in $s_1$
    has to be transformed w.r.t the sequence $s_2 =
    \C{[I(c,0);S(3,c,s)]}$.  The function \C{op\_transform} computes
    such \emph{operational transformations} for an edit sequence $s_1$
    w.r.t another edit sequence $s_2$. A notable transformation rule
    is for a substitution w.r.t another substitution at the same
    position. Since the substituting items could be different, the
    function relies on the item merge function (\C{A.merge}) to merge
    them into a single item. For instance, in Fig.~\ref{fig:list-eg},
    there are simultaneous substitutions to the item \C{c} in $s_1$
    and $s_2$; while $s_1$ substitutes it with \C{d}, $s_2$ does it
    with \C{s}. The merged list therefore substitutes \C{c} with the
    merge of \C{d} and \C{s}, defined as per the merge semantics of
    \C{A.t}.
\end{itemize}
The \C{MList.merge} can now be defined as a function that first
computes edit sequences, $s_1$ and $s_2$, for the two concurrent
lists, $v_1$ and $v_2$ w.r.t their common ancestor $v$, transforms one
of them (say, $s_1$) w.r.t $s_2$ to obtain $s_1'$, and finally applies
(the operations defined by) $s_1'$ to $v_2$ to obtain the merged list.

@@

@@

A non-trivial demonstration illustrating the compositional power of
mergeable types is a shopping list application that allows its users
to collaboratively build a shopping list (listing items in the
decreasing order of their priority), by adding new items with a
specified quantity, deleting existing items, and increasing or
decreasing their quantity. An item (\C{Item.t}) is represented as a
tuple of its name (a string) and the quantity (A \C{Qty.t}).  The
merge function for \C{Item.t} merges the quantities of items (using
\C{Qty.merge}) having the same names. Items with different names are
considered distinct. A shopping list is a mergeable list of items,
i.e., \C{MList(Item)} (although a grocery shopping list may not be
large enough to justify using ropes, \C{MRope(MList(Item))} is
certainly allowed).  \C{MList(Item).merge} automatically lifts the
merge semantics for items to a merge semantics for lists of items.
Fig.~\ref{fig:shopping-list} illustrates its behavior.  Here, two
users, Alice and Bob, concurrently edit a shopping list whose initial
contents are \C{milk} and \C{eggs}. While both simultaneously update
the quantity of eggs, Bob also removes \C{milk} and inserts \C{candy}.
Fig.~\ref{fig:shopping-list-code} shows the \name code for two
concurrent sessions. The merged shopping list contains \C{eggs} and
\C{candy}, where the quantity of eggs is obtained by merging Alice's
\begin{wrapfigure}{r}{.35\textwidth}
  \centering
  \includegraphics[width=0.25\textwidth]{Figures/shopping-list}
  \caption{Merging concurrent shopping lists}
  \label{fig:shopping-list}
  \vspace*{.25in}
\end{wrapfigure}
and Bob's quantity as per the definition of \C{Qty.merge}.

The examples described in this section make certain assumptions about
the concurrency model and the underlying system, such as the existence
of a single lowest common ancestor (LCA) for any pair of versions, the
ability to access a previous version on any branch, and mergeability
of any two concurrent versions. Enforcing these guarantees in a fully
decentralized distributed setting requires addressing non-trivial
theoretical challenges. These, and other details that contribute to
the practicality of our model, such as containing the complexity of
merges, are discussed in subsequent sections.

### 2017-03-15 paper/intro.tex
Fixed many errors.

To illustrate how the kinds of problems described above manifest in
practice, consider how we might write a simple counter library (see
Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two (update)
operations - \C{add} and \C{mult} - that lets a non-negative integer
value be added or multiplied to the counter, resp.  Observe that the
library is written in an idiomatic functional style, with no special
reasoning principles needed to realize desired functionality.  As long
as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instantaneously witnessed on another, which may be in the
process of simultaneously attempting to perform its own \C{Add} or
\C{Mult} action.  While synchronizing the activities of all nodes to
ensure at most one such operation is performed at a time is
impractical, designating a single node to hold the counter state
eschewing replication altogether (as in a typical client-server
configuration), is also an undesirable solution, given the sensitivity
of such architectures to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removing coordination altogether by
simply replicating counter state without having any supporting
consistency protocol is an equally infeasible approach.

A method often adopted to address these concerns is to re-define a
datatype's operations to return \emph{effects} instead of
values~\cite{SPB+11,Burckhardt2014}.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all replicas to
incorporate the effects of the original
operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
with operations re-defined to returns effects.  Observe that the
\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
when applied at a replica (see \C{apply} in the figure), adds \C{x} to
the local counter value.  Note, however, that \C{add} and \C{mult} are
not commutative operations - assuming two replicas have the same
initial counter value of 0, applying the effect of adding 3 and then
multiplying 5 on one replica yields 15, while applying the effect of
first multiplying 5 and then adding 3 yields 3 on the other.  Such
scenarios are possible in a distributed system because there are no
coherence guarantees on the order in which effects are received by
different nodes.  Thus, implementations must be carefully written to
take the lack of commutativity into account when defining how effects
are applied at a replica; in the figure, multiplication is expressed
in terms of addition to avoid the kind of undesirable behavior
described above.

@@

Our main innovation is thus the development of a programming model
that realizes a monadic version control system centered around data,
rather than file, coherence.  The \name (Versioned ML) programming
model comprises the {\sc vst} monad which lets programmers write and
compose concurrent computations around multiple (implicit) versions of
a mergeable datatype, an ordinary ML datatype additionally equipped
with a \emph{merge} function responsible for deriving a consistent
global state from a collection of local versions of that state.  A
computation progresses by forking (i.e., replicating) new versions of
existing versions, allowing synchronization-free local computation to
proceed on those nodes, creating new versions along an existing
\emph{branch} that represents updated local instances, and by merging
branches to realize global consistency.

@@

    \item We describe \name, a programming model that extends ML to
      brings to bear the power and flexibility of a version control
      system to the administration of replicated data. \name hides the
      complexity of version control behind a monad, and exposes its
      functionality via a simple API that lets programmers define and
      compose distributed ML computations around such data.  Issues
      related to replication only manifest implicitly in the
      definition of a merge operation that defines coherence among
      different instances of (presumably) replicated state.

    \item We present a formalization of the \name programming model,
      and establish the conditions under which eventual convergence of
      concurrent version state can be guaranteed.

@@

\noindent The remainder of the paper is organized as follows.  We present the
\name\ programming model informally in the next section.
Sec.~\ref{sec:mergeable_types} develops a series of examples that illustrate
how merge operations can be defined and used. Sec.~\ref{sec:formalization}
defines an operational semantics, and formalizes our intuition of a mergeable
type in terms of morphisms over datatype operations.  We also present soundness
(consistency) and progress guarantees enjoyed by the semantics.
Sec.~\ref{sec:implementation} presents an instantiation of the model suitable
for highly-available distributed environments with network partitions and
failures. Experimental results and benchmarks are given in
Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related} and~\ref{sec:conclusions}
describes related work and conclusions.

>>>>

To illustrate how the kinds of problems described above manifest in
practice, consider how we might write a simple monotonic counter
library (see Fig.~\ref{fig:counter-adt}).  A \C{Counter} supports two
(update) operations - \C{add} and \C{mult} - that lets a non-negative
integer value be added or multiplied to the counter, resp.  Observe
that the library is written in an idiomatic functional style, with no
special reasoning principles needed to realize desired functionality.
As long as applications use the library on a single machine, this
implementation behaves as expected.  However, if the library is used
in the context of a more sophisticated application, say one whose
computation is distributed among a collection of machines, its
behavior can become significantly harder to understand.  In
particular, a distributed implementation might wish to
\emph{replicate} the counter state on each node to improve response
time or fault tolerance.  Unfortunately, adding replication doesn't
come for free.  Attempting to update every replicated copy atomically
is problematic in the absence of sophisticated transaction support,
which impose significant performance penalties.  But, without such
heavyweight mechanisms, applying an \C{Add} operation on one node may
not be instantaneously witnessed on another, which may be in the
process of simultaneously attempting to perform its own \C{Add} or
\C{Mult} action.  While synchronizing the activities of all nodes to
ensure at most one such operation is performed at a time is
impractical, designating a single node to hold the counter state
eschewing replication altogether (as in a typical client-server
configuration), is also an undesirable solution, given the sensitivity
of such architectures to network partitions and server failures, and
the negative performance impact it incurs in geo-distributed
environments~\cite{Walter}.  Removing coordination altogether by
simply replicating counter state without having any supporting
consistency protocol is an equally infeasible approach.

A method often adopted to address these concerns is to re-define a
data type's operations to return \emph{effects} instead of
values~\cite{SPB+11,Burckhardt2014}.  An \emph{effect} is a tag that
identifies the operation to be applied uniformly at all replicas to
incorporate the effects of the original
operation. Fig.~\ref{fig:counter-rdt} shows the \C{Counter} library
with operations re-defined to return effects.  Observe that the
\C{Counter.add x v} operation now returns an \C{Add x} effect, which,
when applied at a replica (see \C{apply} in the figure), adds \C{x} to
the local counter value.  Note, however, that \C{add} and \C{mult} are
not commutative operations - assuming two replicas have the same
initial counter value of 0, applying the effect of adding 3 and then
multiplying 5 on one replica yields 15, while applying the effect of
first multiplying 5 and then adding 3 yields 3 on the other.  Such
scenarios are possible in a distributed system because there are no
coherence guarantees on the order in which effects are received by
different nodes.  Thus, implementations must be carefully written to
take the lack of commutativity into account when defining how effects
are applied at a replica; in the figure, multiplication is expressed
in terms of addition to avoid the kind of undesirable behavior
described above.

@@

Our main innovation is thus the development of a programming model
that realizes a monadic version control system centered around data,
rather than file, coherence.  The \name (Versioned ML) programming
model comprises the \nameMonad (Versioned State) monad which lets
programmers write and compose concurrent computations around multiple
(implicit) versions of a mergeable datatype, an ordinary ML datatype
additionally equipped with a \emph{merge} function responsible for
deriving a consistent global state from a collection of local versions
of that state.  A computation progresses by forking (i.e.,
replicating) new versions of existing versions, allowing
synchronization-free local computation to proceed on those nodes,
creating new versions along an existing \emph{branch} that represents
updated local instances, and by merging branches to realize global
consistency.

@@

    \item We describe \name, a programming model that extends ML to
      bring to bear the power and flexibility of a version control
      system to the administration of replicated data. \name hides the
      complexity of version control behind a monad, and exposes its
      functionality via a simple API that lets programmers define and
      compose distributed ML computations around such data.  Issues
      related to replication only manifest implicitly in the
      definition of a merge operation that defines coherence among
      different instances of (presumably) replicated state.

    \item We present a formalization of the \name programming model,
      and establish its soundess (consistency) and progress guarantees.

@@

\noindent The remainder of the paper is organized as follows.  We
present the \name\ programming model informally in the next section.
Sec.~\ref{sec:mergeable_types} develops a series of examples that
illustrate how merge operations can be defined and composed for
non-trivial data types.  Sec.~\ref{sec:formalization} defines an
operational semantics, and formalizes our intuition of a mergeable
type in terms of morphisms over datatype operations.  We also present
soundness (consistency) and progress guarantees enjoyed by the
semantics.  Sec.~\ref{sec:implementation} presents an instantiation of
the model suitable for highly-available distributed environments with
network partitions and failures. Experimental results and benchmarks
are given in Sec.~\ref{sec:evaluation}.  Secs.~\ref{sec:related}
and~\ref{sec:conclusions} describes related work and conclusions.

### 2017-03-15 paper/meta.tex
Fixed many errors.

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must be the LCA of $v_1$ and $v_2$,
without which \C{merge} may yield unexpected results. This is
demonstrated for the grow-only counter in
Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.6]{Figures/merge-needs-lca}
\caption{This example of a grow-only counter illustrate why \C{merge}
needs a lowest common ancestor, and not just a common ancestor. Both 0
and 2 are common ancestors of 4 and 7, while 2 is their lowest common
ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
11 (incorrect) if 0 is used as the common ancestor for merge, and 9
(correct, because 2+2+3+2 = 9) if 2 is used. }
\label{fig:merge-needs-lca}
\end{wrapfigure}
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.  For example,
in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
by two ``criss-cross'' merges between their respective
branches\footnote{
  When discussing merges and LCAs, we often attribute the properties
  of latest versions on branches to the branches themselves.  For
  instance, when we say two branches merge, in fact their latest
  versions merge. Likewise, LCA of two branches means the LCA of their
  latest versions.
}
resulting in there being two LCAs (5 and 4) for 0 and 3.
Multiple LCAs can occur even without criss-cross merges, as
demonstrated by Fig.~\ref{fig:external-lcas}.
Concurrent versions with multiple LCAs do not lend themselves to
three-way merging. If such versions are latest on their respective
branches, they render the branches unmergeable (since  \C{lca} is no
longer a function) as demonstrated by examples in
Fig.~\ref{fig:many-lcas}. Note that for both the examples in
Fig.~\ref{fig:many-lcas}, no extension of the branching structure
can make the branches merge again. Thus the system is effectively
\emph{partitioned} permanently. This is clearly a problem.

>>>>

When merging two concurrent versions $v_1$ and $v_2$, the common
ancestor argument for \C{merge} must be the LCA of $v_1$ and $v_2$,
without which \C{merge} may yield unexpected results. This is
demonstrated for the monotonic counter in
Fig.~\ref{fig:merge-needs-lca}, where an incorrect count is obtained if a
common ancestor that is not an LCA is used to merge 4 and 7. While in
this example there is a unique LCA for 4 and 7, in general this may
\begin{wrapfigure}{l}{.4\textwidth}
\centering
\includegraphics[scale=0.6]{Figures/merge-needs-lca}
\caption{This example of a monotonic counter illustrate why \C{merge}
needs a lowest common ancestor, and not just a common ancestor. Both 0
and 2 are common ancestors of 4 and 7, while 2 is their lowest common
ancestor (since $0 \preceq 2$). The result (v) of merging 4 and 7 is
11 (incorrect) if 0 is used as the common ancestor for merge, and 9
(correct, because 2+2+3+2 = 9) if 2 is used. }
\label{fig:merge-needs-lca}
\end{wrapfigure}
not be the case. With unrestrained branching and merging, there is no
bound on the number of LCAs a pair of versions can have.  For example,
in Fig.~\ref{fig:criss-cross-lcas}, the merge of 0 with 3 is preceded
by two ``criss-cross'' merges between their respective
branches\footnote{
  When discussing merges and LCAs, we often attribute the properties
  of latest versions on branches to the branches themselves.  For
  instance, when we say two branches merge, in fact their latest
  versions merge. Likewise, LCA of two branches means the LCA of their
  latest versions.
}
resulting in there being two LCAs (5 and 4) for 0 and 3.
Multiple LCAs can occur even without criss-cross merges, as
demonstrated by Fig.~\ref{fig:external-lcas}.
Concurrent versions with multiple LCAs do not lend themselves to
three-way merging. If such versions are latest on their respective
branches, they render the branches unmergeable (since  \C{lca} is no
longer a function) as demonstrated by examples in
Fig.~\ref{fig:many-lcas}. Note that for both the examples in
Fig.~\ref{fig:many-lcas}, no extension of the branching structure
can make the branches merge again. Thus the system is effectively
\emph{partitioned} permanently. This is clearly a problem.
